---
title: "CUNY SPS DATA606 Lab8 - Introduction to linear regression"
author: "Chinedu Onyeka"
date: "October 24th, 2021"
output:
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

The Human Freedom Index is a report that attempts to summarize the idea of "freedom" through a bunch of different variables for many countries around the globe. It serves as a rough objective measure for the relationships between the different types of freedom - whether it's political, religious, economical or personal freedom - and other social and economic circumstances. The Human Freedom Index is an annually co-published report by the Cato Institute, the Fraser Institute, and the Liberales Institute at the Friedrich Naumann Foundation for Freedom.

In this lab, you'll be analyzing data from Human Freedom Index reports from 2008-2016. Your aim will be to summarize a few of the relationships within the data both graphically and numerically in order to find which variables can help tell a story about freedom.

**Getting Started**

Load packages

In this lab, you will explore and visualize the data using the **tidyverse** suite of packages. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
data('hfi', package='openintro')
```


**The data**

The data we're working with is in the openintro package and it's called `hfi`, short for Human Freedom Index.

1.  What are the dimensions of the dataset?  

### **Solution 1: **  

```{r dimension}
dim(hfi)
```

*The dimensions of the dataset is 1458 by 123. That is, there are 1458 observations(rows) and 123 variables(columns).*  

2.  What type of plot would you use to display the relationship between the personal
    freedom score, `pf_score`, and one of the other numerical variables? Plot 
    this relationship using the variable `pf_expression_control` as the predictor.
    Does the relationship look linear? If you knew a country's `pf_expression_control`,
    or its score out of 10, with 0 being the most, of political pressures and 
    controls on media content, would you be comfortable using a linear model to
    predict the personal freedom score?  
    
### **Solution 2: **  

*I will use a scatter plot to display the relationship between personal freedom score `pf_score` and another numerical variable.*  

```{r scatterplot}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +  xlab("Expression Control") + ylab("Freedom Score") +
  labs(title = "Scatterplot of Freedom Score vs Expression Control")
```


*Yes I would be comfortable using a linear model to predict the personal freedom score. The plot shows a linear trend, and other conditions for linear regression appears to be satisfied.*  


If the relationship looks linear, we can quantify the strength of the relationship with the correlation coefficient.

```{r cor}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```

Here, we set the `use` argument to "complete.obs" since there are some observations of NA.

**Sum of squared residuals**  

<div id="boxedtext">
In this section, you will use an interactive function to investigate what we mean by "sum of squared residuals". You will need to run this function in your console, not in your markdown document. Running the function also requires that the `hfi` dataset is loaded in your environment.
</div>

Think back to the way that we described the distribution of a single variable. Recall that we discussed characteristics such as center, spread, and shape. It's also useful to be able to describe the relationship of two numerical variables, such as `pf_expression_control` and `pf_score` above.

3.  Looking at your plot from the previous exercise, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.  
    
### **Solution 3: **  
*There is a strong positive linear relationship between the freedom score (pf_score) and expression control (pf_expression_control). Also, there appears to be some outliers in the plot.*  


Just as you've used the mean and standard deviation to summarize a single variable, you can summarize the relationship between these two variables by finding the line that best follows their association. Use the following interactive function to select the line that you think does the best job of going through the cloud of points.

```{r plotss-expression-score, eval=FALSE}
# This will only work interactively (i.e. will not show in the knitted document)
hfi <- hfi %>% filter(complete.cases(pf_expression_control, pf_score))
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score)
```

After running this command, you'll be prompted to click two points on the plot to define a line. Once you've done that, the line you specified will be shown in black and the residuals in blue. Note that there are 30 residuals, one for each of the 30 observations. Recall that the residuals are the difference between the observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes the sum of squared residuals. To visualize the squared residuals, you can rerun the plot command and add the argument `showSquares = TRUE`.

```{r plotss-expression-score-squares, eval=FALSE}
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the slope and intercept of your line as well as the sum of squares.

4.  Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?  

### **Solution 4: **  
*The smallest sum of squares is 1513.945*  


**The linear model**  

It is rather cumbersome to try to get the correct least squares line, i.e. the line that minimizes the sum of squared residuals, through trial and error. Instead, you can use the `lm` function in R to fit the linear model (a.k.a. regression line).

```{r m1}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)
```

The first argument in the function `lm` is a formula that takes the form `y ~ x`. Here it can be read that we want to make a linear model of `pf_score` as a function of `pf_expression_control`. The second argument specifies that R should look in the `hfi` data frame to find the two variables.

The output of `lm` is an object that contains all of the information we need about the linear model that was just fit. We can access this information using the summary function.

```{r summary-m1}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe the model is shown at the top. After the formula you find the five-number summary of the residuals. The "Coefficients" table shown next is key; its first column displays the linear model's y-intercept and the coefficient of `at_bats`. With this table, we can write down the least squares regression line for the linear model:

\[
  \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control
\]

One last piece of information we will discuss from the summary output is the Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the proportion of variability in the response variable that is explained by the explanatory variable. For this model, 63.42% of the variability in runs is explained by at-bats.

5.  Fit a new model that uses `pf_expression_control` to predict `hf_score`, or
    the total human freedom score. Using the estimates from the R output, write 
    the equation of the regression line. What does the slope tell us in the 
    context of the relationship between human freedom and the amount of political
    pressure on media content?  
    
### **Solution 5: **  

```{r}
lm_hf_score <- lm(hf_score ~ pf_expression_control, data = hfi)
summary(lm_hf_score)
```
slope = 0.349862, intercept = 5.153687; hf_score = 5.153687 + 0.349862*pf_expression_control  
*The slope is the amount by which the human freedom score will increase if the pf_expression_control.*  


**Prediction and prediction errors**

Let's create a scatterplot with the least squares line for `m1` laid on top.

```{r reg-with-line}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

Here, we are literally adding a layer on top of our plot. `geom_smooth` creates the line by fitting a linear model. It can also show us the standard error `se` associated with our line, but we'll suppress that for now.

This line can be used to predict $y$ at any value of $x$. When predictions are made for values of $x$ that are beyond the range of the observed data, it is referred to as *extrapolation* and is not usually recommended. However, predictions made within the range of the data are more reliable. They're also used to compute the residuals.

6.  If someone saw the least squares regression line and not the actual data, how
    would they predict a country's personal freedom school for one with a 6.7 rating
    for `pf_expression_control`? Is this an overestimate or an underestimate, and
    by how much? In other words, what is the residual for this prediction?  
    
### **Solution 6: **  

```{r}
lm_pf_score <- lm(pf_score ~ pf_expression_control, data = hfi)
summary(lm_pf_score)
```
slope = 0.49143, intercept = 4.61707; pf_score = 4.61707 + 0.49143*pf_expression_control  
```{r}
# predict pf_score for a given pf_expression_control rating
pf_expression_control <- 6.7
pf_intercept <- 4.61707
pf_slope <- 0.49143
pf_score_pred <- round((pf_intercept + pf_slope*pf_expression_control),2)
paste0("A country's personal freedom score for a ", pf_expression_control,  
       " rating for pf_expression_control is ", pf_score_pred)
```

```{r}
# residual = actual - predicted
hfi %>% group_by(pf_score) %>% filter(pf_expression_control == 6.7)
# There is no observation with pf_expression_control = 6.7
hfi_samp <- hfi %>% group_by(pf_score) %>% 
  filter(pf_expression_control >= 6.7 & pf_expression_control <= 6.75) %>% 
  select(countries,pf_expression_control, pf_score)
head(hfi_samp, n = 2)
#residual <- pf_score_actual
```

*Looking at how the linear model predicted pf_score for about 2 countries:*
```{r}
pf_intercept <- 4.61707
pf_slope <- 0.49143
pf_exp_ctrl_belize <- hfi_samp$pf_expression_control[1]
pf_exp_ctrl_chile <- hfi_samp$pf_expression_control[2]
pf_score_belize <- hfi_samp$pf_score[1]
pf_score_chile <- hfi_samp$pf_score[2]
pf_score_pred_belize <- round((pf_intercept + pf_slope*pf_exp_ctrl_belize),2)
residual_belize <- round((pf_score_belize - pf_score_pred_belize),3)
paste0("The residual for pf_score for belize is ", residual_belize, " and this is an underestimate.")
pf_score_pred_chile <- round((pf_intercept + pf_slope*pf_exp_ctrl_belize),2)
residual_chile <- round((pf_score_chile - pf_score_pred_chile),3)
paste0("The residual for pf_score for chile is ", residual_chile, " and this is an overestimate.")
```



**Model diagnostics**

To assess whether the linear model is reliable, we need to check for (1) linearity, (2) nearly normal residuals, and (3) constant variability.

**Linearity**: You already checked if the relationship between `pf_score` and `pf_expression_control' is linear using a scatterplot. We should also verify this condition with a plot of the residuals vs. fitted (predicted) values.

```{r residuals}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

Notice here that `m1` can also serve as a data set because stored within it are the fitted values ($\hat{y}$) and the residuals. Also note that we're getting fancy with the code here. After creating the scatterplot on the first layer (first line of code), we overlay a horizontal dashed line at $y = 0$ (to help us check whether residuals are distributed around 0), and we also reanme the axis labels to be more informative.

7.  Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between the two variables?  
    
### **Solution 7: **  

*There is no apparent pattern in the residuals plots and this signifies that there may be a linear relationship between the two variables.*

</br>

**Nearly normal residuals**: To check this condition, we can look at a histogram

```{r hist-res}
ggplot(data = m1, aes(x = .resid)) +
  geom_histogram(binwidth = 0.2) +
  xlab("Residuals")
```

or a normal probability plot of the residuals.

```{r qq-res}
ggplot(data = m1, aes(sample = .resid)) +
  stat_qq()
```

Note that the syntax for making a normal probability plot is a bit different than what you're used to seeing: we set `sample` equal to the residuals instead of `x`, and we set a statistical method `qq`, which stands for "quantile-quantile", another name commonly used for normal probability plots.

8.  Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?  
    
### **Solution 8: **  

*From the histogram and normal plot, it can be seen that the residuals appear to follow a normal model. Hence, the nearly normal residuals condition appear to be met.*

</br>

**Constant variability**:

9.  Based on the residuals vs. fitted plot, does the constant variability condition 
    appear to be met?  

### **Solution 9: **  

*From the residual vs fitted plot, the constant variability condition appears to be met. Based on the conditions for linear regression being satisfied, we can go ahead and use a linear regression to model the relationship.*  


**More Practice**

10.   Choose another freedom variable and a variable you think would strongly
    correlate with it.. Produce a scatterplot of the two variables and fit a 
    linear model. At a glance, does there seem to be a linear relationship?  

### **Solution 10: **  
```{r}
# plot pf_score vs hf_rank
ggplot(data = hfi, aes(x = hf_rank, y = pf_score)) +   geom_point() +
  stat_smooth(method = "lm", se = FALSE) + labs(title = "pf_score vs hf_rank")
```

*At a glance, there seems to be a negative linear relationship between the two variables*


11.   How does this relationship compare to the relationship between 
    `pf_expression_control` and `pf_score`? Use the $R^2$ values from the two 
    model summaries to compare. Does your independent variable seem to predict
    your dependent one better? Why or why not?  
    
### **Solution 11: **  
```{r}
# summary of regression line for pf_score vs pf_expression_control
summary(lm_pf_score)

```

```{r}
lm_pf_score_hf_rank <- lm(pf_score ~ hf_rank, data = hfi)
# summary of regression line for pf_score vs hf_rank
summary(lm_pf_score_hf_rank)
```


*The R-squared for `pf_score` vs `pf_expression control` is 63.42% while the R-squared for `pf_score` vs `hf_rank` is 87.64%. The independent variable in the later seem to predict the dependent variable better because it has a higher R squared value of 87.64%.*  

12.   What's one freedom relationship you were most surprised about and why? Display
    the model diagnostics for the regression model analyzing this relationship.  
    
### **Solution 12: **  
*Surprisingly, `ef_regulation` has a positive relationship with `ef_regulation_business_bribes`. Although the relationship is weak.
```{r}
prob12 <- lm(ef_regulation ~ ef_regulation_business_bribes, data = hfi)
summary(prob12)
```

